---
title: "A primer on structured decision making for invasive species (using dynamic programming)"
author:
  - Olivier Gimenez^1^*

date: '`r Sys.Date()`'

output:
  bookdown::pdf_document2:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    df_print: kable
    keep_tex: yes
    number_sections: no
    toc: no
    includes:
      in_header: header.tex
    
documentclass: article
papersize: a4
pagestyle: plain
geometry: margin=1in
linestretch: 2
fontfamily: mathpazo
fontsize: 11pt

bibliography: 
  - references.bib
  - knitcitations.bib
csl: ecology.csl

link-citations: yes
linkcolor: RoyalBlue
urlcolor: RoyalBlue
links-as-notes: false

---

\linenumbers

\small

^1^ CEFE, Univ Montpellier, CNRS, EPHE, IRD, Montpellier, France

`*` Corresponding author: olivier.gimenez@cefe.cnrs.fr

\normalsize

\vspace{1cm}
\hrule

Invasive species (IAS) pose significant ecological, economic, and public health challenges. Robust methods are needed for guiding effective management. Overall, I hope to demonstrate the flexibility and utility of structured decision making in invasive species management. 

Target journal: soit NeoBiota https://www.reabic.net/journals/mbi/Submission.aspx (1200 euros, plusieurs formats d’article pertinents) ou encore Biological invasions au format INVASION NOTES, ou encore Peer Community journal.

voir NeoBiota,  ; repartir de cet article using structured decision-making to gain consensus on approaches to invasive species control, des articles qu’il cite, et des articles qui l’ont cité ; pour PC Journal ? Faire un petit shiny qui va bien? 

\vspace{3mm}
\hrule
\vspace{5mm}

*Keywords*: Invasive species, Structured decision making, Adaptive management, Stochastic dynamic programming

\newpage

```{r setup, include=FALSE, cache=FALSE, message = FALSE}
library("knitr")
opts_chunk$set(echo = FALSE, warning = TRUE, message = TRUE)
opts_chunk$set(tidy = TRUE, comment = NA, highlight = TRUE)
opts_chunk$set(fig.path = "output/figures/")
```


```{r knitcitations, cache = FALSE}
library(knitcitations)
cleanbib()   
cite_options(citation_format = "pandoc")
```

# Introduction

Invasive species are a significant global issue, with wide-ranging impacts on ecosystems, economies, and public health [@Roy2024; @Petr2020]. 

Effective management of invasive species requires the estimation of population abundance for guiding control efforts and evaluating the success of eradication or regulation programs [@Thompson2021; @Williams2002]. 

Pitch: SDM/AM plusieurs fois recommandées pour management of invasive species. Mais encore peu utilisée. Many reasons. Une d’entre elles est que c’est technique la partie determine trade-offs. Here on participe à l’effort en illustrant la SDP et qqs-unes des variantes (active/passive, POMDP) et dans des cas différents (single pop, metapop, two interacting pops). 

Plusieurs papiers qui préconisent/font de la SDM sur invasive species. Faire une petite review. 

# A primer on structured decision making 

## Why and what

## Stochastic dynamic programming

Example on computer maintenance pour bien expliquer la SDP

# Case studies with coypus

## Single population

Reprendre l'exemple loup de Lulu dans l'article SDP

## Two interacting species

Competition or predator-prey, see SDP-species-interactions/ dans la biblio

## Metapopulation

# Acounting for uncertainty

## POMDP

llustrer POMDP (https://chatgpt.com/share/68814183-9078-8011-9697-c3ee73b27762)

## Adaptive management

Sur une espèce (voir https://chatgpt.com/share/6873c9fd-4df4-8011-97cb-2b6147d505fe ; voir aussi le bouquin de Conroy sur SDM, il y a une figure qui l'explique bien).

# Discussion

## Main messages

## Limitations

## Perspectives

# Acknowledgments

## Data availability statement

Data and code are available at \href{https://github.com/oliviergimenez/counting-rodents}{https://github.com/oliviergimenez/counting-rodents}.

# References

::: {#refs}
:::

